What We Eval in the Shadows

### 2021-04-16 paper/main.tex
Updates

From the \packageNbrunsRnd programs we capture \SEAll side effects from
\SEAllCalls \eval calls in \SEAllSites sites. Again, the challenge is to remove
the accidental side-effects caused by the R virtual machine implementation and
are not related to the user code. For example, the \c{.Random.seed} variable,
which contains the state of the random number generator is saved and restored
from and to the global environment every time the user calls one of the base
routines that use random numbers.
%
% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
% in target environments for side effects.} \label{fig:se-types}
% \end{wrapfigure}
%
Removing them leaves us with \SEUser side effects (\SEUserRatio) from
\SEUserCalls \eval calls (\SEUserCallsRatio) in \SEUserSites sites
(\SEUserSitesRatio). These sites are part of \SEUserFunctions in
\SEUserPackages packages. \SEFunsNighty functions are responsible for 90\% of
side effects. Half of the side-effects come just from three functions:
\c{plyr::allocate\_column} (allocates space for a new data frame column),
\c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object}, which
represents a side-effect that happens in S4 or R6 object environment, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller. The table shows data for both \eval sites and functions. For a function
to have a given target environment class, all of its \eval sites that do side
effects must have the same target environment class The majority of the \eval
sites (\SESitesInOneClass) do all side effects consistently in one environment
class. The same happens on the function level. Almost half of the sites (over a
third of the functions) do side effects in either \emph{Local} or \emph{Object}
environments. This gives a ray of hope for the compiler. Even though it is
possible to do anything anywhere, the data suggest that most side-effects are
sane.

@@

\mypara{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope.
In R, given the support for first-class environments, it can happen anywhere.
The data however suggest that most of side effects from \eval happen in a
predictable environment.

@@

\todo{Aviral's story when he contacted a developer about his use of eval}

>>>>

From the \packageNbrunsRnd programs we capture \SEAllRnd side effects from
\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
side-effects caused by the R virtual machine implementation and are not
related to the user code. For example, the \c{.Random.seed} variable, which
contains the state of the random number generator is saved and restored from
and to the global environment every time the user calls one of the base
routines that use random numbers.
%
% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
% in target environments for side effects.} \label{fig:se-types}
% \end{wrapfigure}
%
Removing them leaves us with \SEUserRnd side effects (\SEUserRatio) from
\SEUserCallsRnd \eval calls (\SEUserCallsRatio) in \SEUserSites sites
(\SEUserSitesRatio). These sites are part of \SEUserFunctions R functions
(\SEUserFunctionsToAllRatio of all functions doing \eval) in \SEUserPackages
packages. From these function, \SEFunsNighty are responsible for 90\% of side
effects. Half of the side-effects come just from three functions:
\c{plyr::allocate\_column} (allocates space for a new data frame column),
\c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object},
which represents a side-effect that happens in S4 or R6 object environment, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller. The table shows data for both \eval sites and functions. For a
function to have a given target environment class, all of its \eval sites that
do side effects must have the same target environment class The majority of the
\eval sites (\SESitesInOneClass) do all side effects consistently in one
environment class. The same happens on the function level. Almost half of the
sites (over a third of the functions) do side effects in either \emph{Local} or
\emph{Object} environments. This gives a ray of hope for the compiler. Even
though it is possible to do anything anywhere, the data suggest that most
side-effects are sane.

@@

\mypara{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope.
In R, given the support for first-class environments, it can happen anywhere
making it \eval more dangerous than it already is. The data however suggest
that first, side-effects from \eval is not as wide spread as in the case of
JavaScript and that over half of them happen in a predictable environment.

@@

% \todo{Aviral's story when he contacted a developer about his use of eval}
